---
title: "PA1_template"
author: "Dave Yachabach"
date: "9/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(ggplot2)
library(dplyr)
library(scales)
```

# PA1 Exercise
## Loading and preprocessing the data

```{r LoadDataset}
filAct <- "activity.csv"
if (!file.exists(filAct)) {
        errorCondition("Ensure Activity.csv exists in the working
                       directory")
        stop("Activity File is Missing")
        }
        
dfAct <- read.csv(filAct)

# Group by day and Sum the steps per day
# 
dfDays <- dfAct %>% group_by(date) %>% 
        summarize(SumSteps = sum(steps,na.rm = TRUE))
```

## What is mean total number of steps taken per day?
### Ignoring missing data



```{r TotalSteps}
# Create histogram of Total Steps per day
g <- ggplot(dfDays)
gh <- g + geom_histogram(aes(SumSteps), binwidth = 2000, color = "blue", fill = "blue", alpha = 0.4)
ghl <- gh + labs(title = "Total Steps Each Day", x = "Total Steps", y = "Frequency (Number of Days)")
ghlm <- ghl + geom_vline(xintercept = mean(dfDays$SumSteps), size = 1, color = "red", alpha = 0.5)
ghlm + geom_vline(xintercept = median(dfDays$SumSteps), size = 1, color = "orange", alpha = 0.5)

# Report Max and Mean
MeanSteps <- round(mean(dfDays$SumSteps))
MedSteps <- median(dfDays$SumSteps)
```

The Mean daily number of steps is `r MeanSteps`.  

The Median daily number of steps is `r MedSteps`.

## What is the average daily activity pattern?

```{r AverageIntervalSteps}
# Build a dataframe of intervals and mean steps
dfIntervals <- dfAct %>% group_by(interval) %>%                            summarise(MeanSteps = mean(steps, na.rm = TRUE))

#Build the line plot
g <- ggplot(dfIntervals, aes(x = interval, y = MeanSteps))
gp <- g + geom_line()
gpl <- gp+ labs(title = "Average Steps in 5-Minute Intervals", 
                x = "5-Minute Interval", y = "Mean Steps")
gpl

# Report the Interval with the max # of mean steps
dfMax <- dfIntervals[which.max(dfIntervals$MeanSteps),]
dfMax
```

The maximum number of mean steps is `r round(dfMax$MeanSteps)` occurring on the 'r dfMax$interval` interval.  

## Imputing missing values

```{r ImputingData}
# Build a dataframe of a count of missing values
dfMissing <- sapply(dfAct, function(x) sum(is.na(x)))

# Build a dataframe with missing data filled with average steps 
# in that interval
dfImpute <- dfAct
dfImpute$steps <- ifelse(is.na(dfImpute$steps) == TRUE, 
                       dfIntervals$MeanSteps[dfIntervals$interval
                        %in% dfImpute$interval], dfImpute$steps)
dfImputeDays <- dfImpute %>% group_by(date) %>% 
        summarize(SumSteps = sum(steps))

# Create histogram of Total Steps per day
ig <- ggplot(dfImputeDays)
igh <- ig + geom_histogram(aes(SumSteps), binwidth = 2000, color = "red",
                           fill = "red", alpha = 0.5)
ighl <- igh + labs(title = "Total Steps Each Day", 
                   subtitle = "Imputed Missing Data",
                   x = "Total Steps", 
                   y = "Frequency (Number of Days)")
ighlm <- ighl + geom_vline(xintercept = mean(dfImputeDays$SumSteps), 
                           size = 1, color = "red", alpha = 0.5)
ighlmv <- ighlm + geom_vline(xintercept = median(dfImputeDays$SumSteps),
                             size = 1, color = "orange", alpha = 0.5)
ighlmv + geom_histogram(data = dfDays, aes(SumSteps), binwidth = 2000, 
                        color = "blue", fill = "blue", alpha = 0.1)

# Report Max and Mean
MeanStepsI <- round(mean(dfImputeDays$SumSteps), digits = 0)
MedStepsI <- round(median(dfImputeDays$SumSteps), digits = 0)
```

After Imputing missing values, the Mean daily number of steps is `r MeanStepsI` as compared to `r MeanSteps` with missing values.    

After Imputing missing values, the Median daily number of steps is `r MedStepsI` as compared to `r MedSteps` with missing values.  

## Are there differences in activity patterns between weekdays and weekends?

```{r Weekend}
# Build dataframe with an additional column describing "weekend" or 
# "weekday
# 
dfDayOfWk <- dfImpute %>% mutate(WkDay = weekdays(as.Date(dfImpute$date)))

# Change name of the day of the week to "weekday" or "weekend" and make it 
# a factor.
dfDayOfWk$WkDay <- ifelse(dfDayOfWk$WkDay %in% c("Saturday", "Sunday"), 
                          "weekend", "weekday")
dfDayOfWk$WkDay <- factor(dfDayOfWk$WkDay)

# Get the Intervals and Mean per day by weekday value
dfIntWkDay <- dfDayOfWk %>% group_by(interval,WkDay) %>% 
        summarise(MeanSteps = mean(steps))

# Now plot the result
# 
g <- ggplot(dfIntWkDay)
gp <- g + geom_line(aes(x=interval, y = MeanSteps, group = WkDay)) + 
        facet_grid(WkDay~.)
gp + labs(title = "Activity During Weekday vs Weekend", 
          y = "Mean Number of Steps", x = "5-Minute Interval")
```

